\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2020-06-18 sections/abstract.tex
Updates to draft

\begin{abstract}
  {
    \noindent
    Microbes tend to organize into communities consisting of hundreds of species involved in complex interactions with each other.
    16S ribosomal RNA (16S rRNA) amplicon profiling provides snapshots that reveal the phylogenies and abundance profiles of these microbial communities.
    These snapshots, when collected from multiple samples, have the potential to reveal which microbes co-occur, providing a glimpse into the network of influences in these communities.
    The inference of networks from 16S data is prone to statistical artifacts.
    There are many tools for performing each step of the 16S analysis workflow, but the extent to which these steps affect the final network is still unclear.
    In this study, we perform a meticulous analysis of each step of a pipeline that processes 16S sequencing data into a network of microbial associations.
    Through this process, we map how different choices of algorithms and parameters affect the co-occurrence network, and estimate steps which contribute most significantly to the variance. We further determine the tools and parameters that generate the most accurate and robust co-occurrence networks, based on comparison with mock and synthetic datasets.
    Ultimately, we develop a standardized pipeline that follows these default tools and parameters, but that can also help explore the outcome of any other combination of choices.
    We envisage that this pipeline for processing 16S sequencing data into networks of microbial co-occurrences could be used for integrating multiple data-sets, and for generating comparative analyses and consensus networks useful for detecting disease-related patterns.
    % TODO: Move package to segrelab github and change name to MiCoNE
    The pipeline is available as a Python package at \href{https://github.com/dileep-kishore/mindpipe}{https://github.com/dileep-kishore/mindpipe}
  }
\end{abstract}

% Insert keywords here
\keywords{Microbiome, 16S rRNA, Pipeline, Interaction, Network, Correlations, Qiime, Co-occurrence}

>>>>

\begin{abstract}
  {
    \noindent
    Microbes tend to organize into communities consisting of hundreds of species involved in complex interactions with each other.
    16S ribosomal RNA (16S rRNA) amplicon profiling provides snapshots that reveal the phylogenies and abundance profiles of these microbial communities.
    These snapshots, when collected from multiple samples, have the potential to reveal which microbes co-occur, providing a glimpse into the network of influences in these communities.
    The inference of networks from 16S data is prone to statistical artifacts.
    There are many tools for performing each step of the 16S analysis workflow, but the extent to which these steps affect the final network is still unclear.
    In this study, we perform a meticulous analysis of each step of a pipeline that processes 16S sequencing data into a network of microbial associations.
    Through this process, we map how different choices of algorithms and parameters affect the co-occurrence network, and estimate steps which contribute most significantly to the variance. We further determine the tools and parameters that generate the most accurate and robust co-occurrence networks, based on comparison with mock and synthetic datasets.
    Ultimately, we develop a standardized pipeline that follows these default tools and parameters, but that can also help explore the outcome of any other combination of choices.
    We envisage that this pipeline for processing 16S sequencing data into networks of microbial co-occurrences could be used for integrating multiple data-sets, and for generating comparative analyses and consensus networks useful for detecting disease-related patterns.
        The pipeline is available as a Python package at \href{https://github.com/segrelab/MiCoNE}{https://github.com/segrelab/MiCoNE}
  }
\end{abstract}

% Insert keywords here
\keywords{Microbiome, 16S rRNA, Pipeline, Interaction, Denoising, Taxonomy, Network Inference, Correlations, Qiime, Co-occurrence, Networks}

### 2020-06-18 sections/discussion.tex
Updates to draft

The choice of taxonomy database was found to be the most important factor in the inference of a microbial co-occurrence network, contributing $\sim20\%$ of the total variance.
The frequent changes in the taxonomy nomenclature coupled with the frequency of updates to the various 16S reference databases create inherent differences \cite{Balvociute2017} in taxonomy hierarchies in these databases.
Our analysis did not reveal one particular reference database to perform better in all scenarios. We suggest that that choice of the database should be made based on possible reported or inferred biases in the representation of given biomes in a specific databases \cite{Balvociute2017}.
The default reference database in the pipeline is the NCBI 16S RefSeq database as it is more frequently updated and is most compatible with the blast+ query tool.
We also enable users to use custom databases \cite{Griffen2011,Ritari2015} with the blast and naive bayes classifiers that are incorporated into the pipeline (from \ac{qiime2}).

@@

% TODO: Uses in MIND, exploration, standardization, reproducibility

>>>>

The choice of taxonomy database was found to be the most important factor in the inference of a microbial co-occurrence network, contributing $\sim20\%$ of the total variance.
The frequent changes in the taxonomy nomenclature coupled with the frequency of updates to the various 16S reference databases create inherent differences \cite{Balvociute2017} in taxonomy hierarchies in these databases.
Our analysis did not reveal one particular reference database to perform better in all scenarios. We suggest that that choice of the database should be made based on possible reported or inferred biases in the representation of given biomes in a specific databases \cite{Balvociute2017}.
The default reference database in the pipeline is the \ac{ncbi} 16S RefSeq database as it is more frequently updated and is most compatible with the blast+ query tool.
We also enable users to use custom databases \cite{Griffen2011,Ritari2015} with the blast and naive bayes classifiers that are incorporated into the pipeline (from \ac{qiime2}).

@@

% TODO: Uses in MIND (?)

### 2020-06-18 sections/introduction.tex
Updates to draft

  High-throughput metagenomic sequencing techniques (such as 16S rRNA gene profiling) help detect, identify and quantify a large part of the constitutive microorganisms of a microbiome \cite{Jovel2016}.
  Knowledge of the community composition at a particular instance in time would enable us to derive partial insights into the underlying dynamics.
  With the advancement in DNA sequencing technologies \cite{Narihiro2017} and data processing methods much more information can be extracted from these microbial community samples than ever before.
  These advances have led to large-scale data collection efforts involving environmental (Earth Microbiome Project) \cite{Thompson2017}, marine (Tara Oceans Project) \cite{Zhang2015} and human-associated microbiota (Human Microbiome Project) \cite{HumanMicrobiomeProjectConsortium2012} and thus led the generation of a wealth of data about these microbial communities.

@@

  One of the limitations of this type of analysis is that 16S sequencing data is limited by many factors such as resolution, sequencing depth, compositional nature, sequencing errors and copy number variations.
  Many different analysis methods have been developed \cite{Callahan2016,Amir2017,Friedman2012,Kurtz2015} to counter these limitations and remove statistical artifacts, leading to the existence of a myriad of tools for each step of the data analysis workflow.
  The different methods and tools developed to solve these issues infer vastly different community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making it difficult to reliably compare networks across different publications and studies.
  Conversely, given the lack of comprehensive comparisons between directly observed microbial interactions (e.g. from co-culture experiments) and co-occurence networks, there is no straightforward way to determine which set of tools or methods generate the most accurate networks.

 Previous work has developed popular platforms (like MG-RAST \cite{Keegan2016}, Qiita \cite{qiita}) and tools (such as QIIME \cite{Caporaso2010}) that provide pipelines for 16S data analysis. However, existing tools typically are not focused on the effects of upstream statistical methods on the inferred co-occurrence networks. Furthermore, no organized framework currently exist to systematically analyze and compare existing components of the data-analysis from amplicons to networks.

  In this study, we present a standardized 16S data-analysis pipeline called \ac{micone} that produces robust and reproducible co-occurrence networks from community 16S sequence data, and allow users to interactively explore how the network would change upon using different alternative tools and parameters at each step.
  % TODO: Link or describe MIND here
  Our pipeline is coupled to an online integrative tool for the organization, visualization and analysis of inter-microbial networks.
  In addition to making this tool freely available, we implemented a systematic comparative analysis to determine which steps of the pipeline have the largest influence on the final network, and what choice seems to provide best agreement with the tested mock and synthetic datasets.
  We believe that these steps will ensure better reproducibility and easier comparison of co-occurrence networks across data-sets.
  We expect that our tool will also be useful for benchmarking future alternative methods, and for ensuring a transparent evaluation of the possible biases introduced by the use of specific tools.

>>>>

  High-throughput metagenomic sequencing techniques (such as 16S rRNA gene profiling) help detect, identify and quantify a large part of the constitutive microorganisms of a microbiome \cite{Jovel2016}.
  Knowledge of the community composition at a particular instance in time would enable us to derive partial insights into the underlying dynamics.
  With the advancement in DNA sequencing technologies \cite{Narihiro2017} and data processing methods much more information can be extracted from these microbial community samples than ever before.
  These advances have led to large-scale data collection efforts involving environmental (\acl{emp}) \cite{Thompson2017}, marine (Tara Oceans Project) \cite{Zhang2015} and human-associated microbiota (Human Microbiome Project) \cite{HumanMicrobiomeProjectConsortium2012} and thus led the generation of a wealth of data about these microbial communities.

@@

  One of the limitations of this type of analysis is that 16S sequencing data is limited by many factors such as resolution, sequencing depth, compositional nature, sequencing errors and copy number variations.
  Many different analysis methods have been developed \cite{Callahan2016,Amir2017,Friedman2012,Kurtz2015} to counter these limitations and remove statistical artifacts, leading to the existence of a myriad of tools for each step of the data analysis workflow.
  The different methods and tools developed to solve these issues infer vastly different community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making it difficult to reliably compare networks across different publications and studies.
  Conversely, given the lack of comprehensive comparisons between directly observed microbial interactions (e.g. from co-culture experiments) and co-occurrence networks, there is no straightforward way to determine which set of tools or methods generate the most accurate networks.

 Previous work has developed popular platforms (like MG-RAST \cite{Keegan2016}, Qiita \cite{qiita}) and tools (such as QIIME \cite{Caporaso2010}) that provide pipelines for 16S data analysis.
 However, existing tools typically are focused on \ac{otu} generation and not on the effects of upstream statistical methods on the inferred co-occurrence networks.
 Furthermore, no organized framework currently exist to systematically analyze and compare existing components of the data-analysis from amplicons to networks.

  In this study, we present a standardized 16S data-analysis pipeline called \ac{micone} that produces robust and reproducible co-occurrence networks from community 16S sequence data, and allow users to interactively explore how the network would change upon using different alternative tools and parameters at each step.
  % TODO: Link or describe MIND here (?)
  Our pipeline is coupled to an online integrative tool for the organization, visualization and analysis of inter-microbial networks.
  In addition to making this tool freely available, we implemented a systematic comparative analysis to determine which steps of the pipeline have the largest influence on the final network, and what choice seems to provide best agreement with the tested mock and synthetic datasets.
  We believe that these steps will ensure better reproducibility and easier comparison of co-occurrence networks across data-sets.
  We expect that our tool will also be useful for benchmarking future alternative methods, and for ensuring a transparent evaluation of the possible biases introduced by the use of specific tools.

### 2020-06-18 sections/methods.tex
Updates to draft

  \vspace{-5mm}
  The study uses three kinds of 16S rRNA sequencing datasets: real datasets, mock datasets and synthetic datasets.
  Real datasets are collections of sequencing reads obtained from naturally occurring microbial community samples.
  The current study used healthy stool samples from a fecal microbiome transplant study \cite{Kang2017} and healthy saliva samples from a periodontal disease study~\cite{Chen2018} as real datasets for analysis.
  The mock community 16S datasets are real sequencing data obtained for artificially assembled collections of species in known proportions. The mock datasets used for this study, obtained from mockrobiota~\cite{Bokulich2016}, were labelled mock4, mock12 and mock16.
  The mock4 community is composed of 21 bacterial strains. Two replicate samples from mock4  contain all species in equal abundances, and two additional replicate samples contain the same species in unequal abundances.
  The mock12 community is composed of 27 bacterial strains that include closely related taxa with some pairs having only one to two nucleotide difference from another.
  The mock16 community is composed of 49 bacteria and 10 archea, all represented in equal amount.
  The synthetic datasets were generated using an artificial read simulator called ART~\cite{Huang2012}.
  Three different microbial composition profiles were used as input; reads were generated using a soil and water microbiome composition profiles from the Earth Microbiome Project~\cite{Thompson2017} and healthy gut microbiome project from the Human Microbiome Project~\cite{HumanMicrobiomeProjectConsortium2012}.
  The reads are simulated using the NCBI RefSeq database as the reference sequence pool and the "art\_illumina" sequence profile with a mutation rate of 2\%.
  The scripts used to generate the synthetic data are in the scripts folder of the respository (\href{https://github.com/dileep-kishore/mind-pipeline-paper}{https://github.com/dileep-kishore/mind-pipeline-paper}).

@@

  \subsection*{Denoising and Clustering (DC)}
  \vspace{-5mm}
  This module deals with processing the raw 16S sequence data into \ac{otu} or \ac{esv} count tables.
  It consists of the following processes: quality control, denoising (or clustering) and chimera checking.
  The quality control process handles the demultiplexing and quality control steps such as trimming adapters and trimming low-quality nucleotide stretches from the sequences.
  The denoise/cluster process handles the conversion of the demultiplexed, trimmed sequences into \ac{otu} or \ac{esv} count tables (some methods, like closed reference and open reference clustering, perform clustering and taxonomy assignment in the same step).
  The chimera checking process handles the removal of chimeric sequences created during the \ac{pcr} step.
  The output of this module is a matrix of counts, that describes the number of reads of a particular \ac{otu} or \ac{esv} (rows of the matrix) present in each sample (columns of the matrix).
  The options currently available in the pipeline for denoising and clustering: open reference clustering, closed reference clustering and de novo clustering methods from \ac{qiime1}~\cite{Caporaso2010} and denoising methods from \ac{dada2}~\cite{Callahan2016} and Deblur~\cite{Amir2017}.
  The quality filtering and chimera checking tools are derived from those used in \ac{qiime2} and \ac{dada2}.

  \subsection*{Taxonomy Assignment (TA)}
  \vspace{-5mm}
  This module deals with assigning taxonomies to either the representative sequences of the \ac{otu}s or directly to the \ac{esv}s.
  In order to assign taxonomies to a particular sequence we need a taxonomy database and a query tool.
  The taxonomy database contains the collection of 16S sequences of micro-organisms of interest and the query tool allows one to compare a sequence of interest to all the sequences in the database to identify the best matches. Finally, a consensus method is used to identify the most probable match from the list of best matches.
  The pipeline incorporates \ac{gg}~\cite{DeSantis2006}, SILVA~\cite{Quast2012} and the \ac{ncbi}~\cite{Sayers2009} databases for taxonomy assignment and the Naive Bayes classifier from \ac{qiime2} and \ac{ncbi} blast as the query tools (from \ac{qiime2}).
  The consensus algorithm used is the default method used by the classifiers in \ac{qiime2}.

  % TODO: Add references and basic equations
  \subsection*{OTU and ESV Processing (OP)}
  \vspace{-5mm}
  This module deals with normalization, filtering and applying transformations to the \ac{otu} or \ac{esv} counts matrix.
  Rarefaction is a normalization technique used to overcome the bias that might arise due to variable sampling depth in different samples. This is performed either by sub-sampling or by normalization of the matrix to the lowest sampling depth.
  Rarefaction is usually followed by filtering, which is performed to remove samples or features (\ac{otu}s or \ac{esv}s) from the count matrix that are sparse. In order to determine the filtering threshold we fix the number of samples and correlation detection power needed and determine the number of features to be used.
  Finally, transformations are performed in order to correct for and overcome the compositional bias that is inherent in a counts matrix (in most cases this is handled by the network inference algorithm).

  % TODO: Add reference for bootstrapping
  \subsection*{Network Inference (NI)}
  \vspace{-5mm}
  This module deals with the inference of co-occurrence associations from the \ac{otu} or \ac{esv} counts matrix.
  These associations can be represented as a network, with nodes representing taxonomies of the micro-organisms and edges representing the association between them.
  A null model is created by re-sampling and bootstrapping the correlation/interaction matrix and is used to calculate the significance of the inferred associations by calculating the p-values against this null model.
  The pipeline includes Pearson, Spearman and \ac{sparcc}~\cite{Friedman2012} as the pairwise correlation metrics, and \ac{spieceasi}~\cite{Kurtz2015}, \ac{mldm}~\cite{Yang2017} and \ac{magma} \hl{in biorxiv [ADD REF]} as the direct association metrics.
  The empirical Browns method \hl{ref ADD REF} is used for combining p-values from the various methods to obtain a consensus p-value, which is used to create the consensus network.

@@

In other words, $N$ is the merged table and each column $N_i$ is the vector representation of one of the networks.

We next use linear regression to express each network Ni as a linear function of categorical variables that describe the possible options in each of the first three steps of the pipeline.

  % TODO: Explain the categorical linear model and ANOVA better

In particular, we infer parameters $\alpha_i$ such that:
   \begin{equation*}
       N_i = \sum_{j=1}^5 \left( \alpha^{DC(j)}_i.\delta^{DC(j)}_i \right) +
             \sum_{j=1}^3 \left( \alpha^{TA(j)}_i.\delta^{TA(j)}_i \right) +
             \sum_{j=1}^2 \left( \alpha^{OP(j)}_i.\delta^{OP(j)}_i \right) +
             \epsilon_i
   \end{equation*}

   $N_i = \mathbf{\alpha_i^{DC}} \cdot \mathbf{ \delta_i^{DC}}$

@@

  \subsection*{Code and Data Availability}
  pipeline: \href{}{} \\
  data and scripts related to the paper: \href{}{}

>>>>

  \vspace{-5mm}
  The study uses three kinds of 16S rRNA sequencing datasets: real datasets, mock datasets and synthetic datasets.
  Real datasets are collections of sequencing reads obtained from naturally occurring microbial community samples.
  The current study used healthy stool samples from a fecal microbiome transplant study~\cite{Kang2017} and healthy saliva samples from a periodontal disease study~\cite{Chen2018} as real datasets for analysis.
  The mock community 16S datasets are real sequencing data obtained for artificially assembled collections of species in known proportions.
  The mock datasets used for this study, obtained from mockrobiota~\cite{Bokulich2016}, were labelled mock4, mock12 and mock16.
  The mock4 community is composed of 21 bacterial strains.
  Two replicate samples from mock4  contain all species in equal abundances, and two additional replicate samples contain the same species in unequal abundances.
  The mock12 community is composed of 27 bacterial strains that include closely related taxa with some pairs having only one to two nucleotide difference from another.
  The mock16 community is composed of 49 bacteria and 10 archea, all represented in equal amount.
  The synthetic datasets were generated using an artificial read simulator called ART~\cite{Huang2012}.
  Three different microbial composition profiles were used as input; reads were generated using a soil and water microbiome composition profiles from the \ac{emp}~\cite{Thompson2017} and healthy gut microbiome project from the fecal microbiome transplant study~\cite{Kang2017}.
  The reads are simulated using the NCBI RefSeq database as the reference sequence pool and the "art\_illumina" sequence profile with a mutation rate of 2\%.
  The scripts used to generate the synthetic data are in the scripts folder of the respository (\href{https://github.com/segrelab/mind-pipeline-paper}{https://github.com/segrelab/mind-pipeline-paper}).

@@

  \subsection*{Denoising and Clustering (DC)}
  \vspace{-5mm}
  This module deals with processing the raw 16S sequence data into \ac{otu} or \ac{esv} count tables.
  It consists of the following processes: quality control, denoising (or clustering) and chimera checking.
  The quality control process handles the demultiplexing and quality control steps such as trimming adapters and trimming low-quality nucleotide stretches from the sequences.
  The denoise/cluster process handles the conversion of the demultiplexed, trimmed sequences into \ac{otu} or \ac{esv} count tables (some methods, like closed reference and open reference clustering, perform clustering and taxonomy assignment in the same step).
  The chimera checking process handles the removal of chimeric sequences created during the \ac{pcr} step.
  The output of this module is a matrix of counts, that describes the number of reads of a particular \ac{otu} or \ac{esv} (rows of the matrix) present in each sample (columns of the matrix).
  The options currently available in the pipeline for denoising and clustering: open reference clustering, closed reference clustering and de novo clustering methods from \ac{qiime1} v1.9.1~\cite{Caporaso2010} and denoising methods from \ac{dada2} v1.14~\cite{Callahan2016} and Deblur v1.1.0~\cite{Amir2017}.
  The quality filtering and chimera checking tools are derived from those used in \ac{qiime2} v2019.10.0 and \ac{dada2}.

  \subsection*{Taxonomy Assignment (TA)}
  \vspace{-5mm}
  This module deals with assigning taxonomies to either the representative sequences of the \ac{otu}s or directly to the \ac{esv}s.
  In order to assign taxonomies to a particular sequence we need a taxonomy database and a query tool.
  The taxonomy database contains the collection of 16S sequences of micro-organisms of interest and the query tool allows one to compare a sequence of interest to all the sequences in the database to identify the best matches.
  Finally, a consensus method is used to identify the most probable match from the list of best matches.
  The pipeline incorporates \ac{gg} 13\_8~\cite{DeSantis2006}, SILVA 132~\cite{Quast2012} and the \ac{ncbi} (16S RefSeq as of Oct 2019)~\cite{Sayers2009} databases for taxonomy assignment and the Naive Bayes classifier from \ac{qiime2} and \ac{ncbi} blast as the query tools (from \ac{qiime2}).
  The consensus algorithm used is the default method used by the classifiers in \ac{qiime2}.

  % TODO: Add references and basic equations
  \subsection*{OTU and ESV Processing (OP)}
  \vspace{-5mm}
  This module deals with normalization, filtering and applying transformations to the \ac{otu} or \ac{esv} counts matrix.
  Rarefaction is a normalization technique used to overcome the bias that might arise due to variable sampling depth in different samples.
  This is performed either by sub-sampling or by normalization of the matrix to the lowest sampling depth \cite{Weiss2015}.
  Rarefaction is usually followed by filtering, which is performed to remove samples or features (\ac{otu}s or \ac{esv}s) from the count matrix that are sparse.
  In order to determine the filtering threshold we fix the number of samples and correlation detection power needed and determine the number of features to be used.
  Finally, transformations are performed in order to correct for and overcome the compositional bias that is inherent in a counts matrix (in most cases this is handled by the network inference algorithm).

  \subsection*{Network Inference (NI)}
  \vspace{-5mm}
  This module deals with the inference of co-occurrence associations from the \ac{otu} or \ac{esv} counts matrix.
  These associations can be represented as a network, with nodes representing taxonomies of the micro-organisms and edges representing the association between them.
  A null model is created by re-sampling and bootstrapping the correlation/interaction matrix and is used to calculate the significance of the inferred associations by calculating the p-values against this null model \cite{Watts2018}.
  The pipeline includes Pearson, Spearman and FastSpar v0.0.10 (a faster implementation of \ac{sparcc})~\cite{Watts2018} as the pairwise correlation metrics, and \ac{spieceasi} v1.0.7~\cite{Kurtz2015}, \ac{mldm} v1.1~\cite{Yang2017} and \ac{magma}~\cite{Cougoul2019} as the direct association metrics.
  The empirical Browns method is used for combining p-values from the various methods to obtain a consensus p-value, which is used to create the consensus network.

@@

  In other words, $N$ is the merged table and each column $N_i$ is the vector representation of one of the networks and each row $L_i$ represents the one particular edge in all networks (assigned 0 if the edge does not exist in the network).

  We next use linear regression to express each link $L_i$ as a linear function of categorical variables that describe the possible options in each of the first three steps of the pipeline.

  % TODO: Explain the categorical linear model and ANOVA better
  In particular, we infer parameters $\alpha_i$ such that:
   \begin{equation*}
       L_i = \sum_{j=1}^5 \left( \alpha^{DC(j)}_i.\delta^{DC(j)}_i \right) +
             \sum_{j=1}^3 \left( \alpha^{TA(j)}_i.\delta^{TA(j)}_i \right) +
             \sum_{j=1}^2 \left( \alpha^{OP(j)}_i.\delta^{OP(j)}_i \right) +
             \epsilon_i
   \end{equation*}

@@

  \subsection*{Code and Data Availability}
  pipeline: \href{https://github.com/segrelab/MiCoNE}{https://github.com/segrelab/MiCoNE} \\
  data and scripts related to the paper: \href{}{}
